Cycle Route Uptake and Scenario Estimation (CRUSE): an open and nationally scalable approach for strategic cycle network planning tool development \\\large{A case study of Ireland}

### 2024-03-23 paper.qmd
Add content from @eugenividal

# Methods and data {#sec-methods}

The methods used to generate the evidence presented in the CRUSE Tool build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionized strategic cycle planning in England and Wales" by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[@lovelace2023].

@@

Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui.

>>>>

# Tools for estimating cycling potential

The tools for estimating geographical distribution of cycling potential have advanced substantially in recent years. Developed by both researchers and professionals, they have evolved from simple area-based static models tailored to specific regions to more complex generalisable open-source online maps at several levels of analysis open to the public. Table 1 summarises the most relevant.

Parkin et al. (2007) presented a regression model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level. They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.

Transport for London (2010) created the Analysis of Cycling Potential (ACP) tool to identify current trips that could be cycled instead of using mechanised modes. The tool is based on a mode choice model that assesses the likelihood of trips being cycled and uses data from the TfL London Travel Demand Survey as its primary input dataset. The tool generates tables, graphs, and heatmaps showing the potential for cycling for different trips purposes in London. This information has been used to guide local cycling initiatives, such as determining locations for new hire stations. The tool, however, it does not use origin-destination data, which makes it less useful for route-based interventions. An updated analysis was conducted in 2017 (Transport for London, 2016).

Larsen et al. (2013) designed a tool to prioritize the placement of new cycling facilities in Montreal, Canada. The tool used survey, vehicle collision, and origin-destination data. To create the index, they first identified a set of relevant indicators for prioritising locations for infrastructure investment. Then, they imposed a grid and aggregated the indicators spatially, associating them with grid’s cells. Finally, they combined the aggregated grid cell indicators into the prioritization index. The tool generated heatmaps recommending cycling infrastructure upgrades in specific locations.

Zhang et al. (2014) used a regression model to examine the effect of socioeconomic factors on willingness to cycle in Belo Horizonte City, Brazil, while considering individual heterogeneity. To develop the model, they employed a bicycle use survey conducted in Belo Horizonte. The results indicated a positive association between willingness to cycle and middle income, people renting apartments, and higher education, and a negatively association with distance to work and car use. Subsequently, using the calibrated model along with census and OD data, they estimated the willingness to commute cycling for the entire city.

Phillips and Range (2017) created the Cycling Potential Tool (CPT) for Scotland to provide evidence-based insights for decision-making regarding local cycling infrastructure in Scotland. The CPT is a GIS toolbox consisting of two main modules: the base environmental module and the quality of service module. The base environmental module is based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) and produces heatmaps for each factor as well as a combined score. The quality of service module evaluates the network within each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).

The Propensity to Cycle Tool (2017) was initially based on current and future potential uptake of cycling for single stage travel to work at desire line, zone, route, and route network levels (Lovelace et al. 2017). Extensions of the PCT approach have included estimation of benefits at the individual level (Woodcock et al., 2018), addition of travel to school network (Goodman et al., 2019), and improved modeling of impacts on health, environmental and distributional outcomes (Woodcock et al., 2021). Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018. The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal. In Portugal the ‘biclaR’ project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region. The resulting evidence is available in an interactive web application hosted at biclar.tmlmobilidade.pt (Félix et al., 2022). biclaR includes estimates of impacts, using the World Health Organisation (WHO) ‘HEAT for Cycling’ tool and an ‘intermodality’ scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data. An important feature of the PCT is that it is open source and publicly available (at www.pct.bike), allowing its use by all stakeholders in the transport planning process.

Silva et al. (2021)  designed the Gross Potential for Cycling (GPC) tool to identify key areas for the implementation of cycling infrastructure and other cycling measures. The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance). These indicators are calculated considering topography, road hierarchy, and average congestion and presented on a scale from 5 to 10. They are then combined into an overall score, weighted by their impact on cycling according to the literature. The GPC is publicly available (at https://boost.up.pt/en/ferramentas/gpc) and is specially designed for local council-level practitioners. Its practical value has been assessed through a series of experiential workshops involving planning practitioners from various Portuguese contexts (Silva et al., 2022).

The methods used to generate the evidence present in the CRUSE Tool, described in the next section, build on the PCT and seek to overcome three of the PCT methodological limitations: 1) low resolution of data, with routes starting and ending in administrative zone centroids, 2) limited coverage of trip purposes beyond travel to work and school, and 3) a web interface that was not user-friendly or intuitive.

# Methods and data {#sec-methods}
<!-- 
The methods used to generate the evidence presented in the CRUSE Tool build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionized strategic cycle planning in England and Wales" by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[@lovelace2023].

@@

Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui. -->

### 2024-03-23 paper.tex
Add content from @eugenividal

@@

The methods used to generate the evidence presented in the CRUSE Tool
build on the Propensity to Cycle Tool (PCT), which was originally funded
by the UK's Department for Transport and developed by a multi-university
team. An important feature of the PCT is that it is open source and
publicly available (at \href{https://www.pct.bike/}{www.pct.bike}),
allowing its use by all stakeholders in the transport planning process
\citep{lovelace2017}. The PCT approach has had major policy impacts, as
outlined in Research Excellence Framework (REF) impact case studies,
which demonstrate that the tool ``revolutionized strategic cycle
planning in England and Wales'' by overcoming the barriers to cycling
investment imposed by lack of evidence on cycling
potential\citep{lovelace2023}.

The first version of the PCT was based on current and future potential
uptake of cycling for single stage \emph{travel to work} at desire line,
zone, route, and route network levels \citep{lovelace2016}. It was
launched in April 2017 as the government-endorsed tool for strategic
cycle network planning and part of the Cycling and Walking Investment
Strategy \citep{cycling2017}. Extensions of the PCT approach have
included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modeling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

The PCT approach has been applied in other countries, including Ireland
(the topic of this paper), Scotland, and Portugal. In Portugal the
`biclaR' project, based on methods underlying the PCT, has been
developed and deployed for the Lisbon metro region. The resulting
evidence is available in an interactive web application hosted at
\href{https://biclar.tmlmobilidade.pt}{biclar.tmlmobilidade.pt}
\citep{felix2023}. biclaR includes estimates of impacts, using the World
Health Organisation (WHO) `HEAT for Cycling' tool and an `intermodality'
scenario that combines cycling with currently available public transit
options based on General Transit Feed Specification (GTFS) data.

The CRUSE Tool seeks to overcome the following methodological
limitations of the original PCT:

\begin{itemize}
\tightlist
\item
  Low resolution of data, with routes starting and ending in
  administrative zone centroids
\item
  Limited coverage of trip purposes beyond travel to work and school
\item
  A web interface that was not user-friendly or intuitive
\end{itemize}

Methods were developed to overcome each of these, as outlined in
Section~\ref{sec-trip-purposes} to Section~\ref{sec-ui}.

>>>>

\section{Tools for estimating cycling
potential}\label{tools-for-estimating-cycling-potential}

The tools for estimating geographical distribution of cycling potential
have advanced substantially in recent years. Developed by both
researchers and professionals, they have evolved from simple area-based
static models tailored to specific regions to more complex generalisable
open-source online maps at several levels of analysis open to the
public. Table 1 summarises the most relevant.

Parkin et al.~(2007) presented a regression model to estimate the
percentage of cycling trips to work in England and Wales based on
socioeconomic, transportation, and physical factors at the small area
(wards) level. They found that areas with a higher percentage of
females, non-whites, car ownership, lower socioeconomic classes, income,
distance to work, population density, poor highway conditions, hills,
rainfall, and fewer off-road bicycle routes tended to have a lower
proportion of cycling to work.

Transport for London (2010) created the Analysis of Cycling Potential
(ACP) tool to identify current trips that could be cycled instead of
using mechanised modes. The tool is based on a mode choice model that
assesses the likelihood of trips being cycled and uses data from the TfL
London Travel Demand Survey as its primary input dataset. The tool
generates tables, graphs, and heatmaps showing the potential for cycling
for different trips purposes in London. This information has been used
to guide local cycling initiatives, such as determining locations for
new hire stations. The tool, however, it does not use origin-destination
data, which makes it less useful for route-based interventions. An
updated analysis was conducted in 2017 (Transport for London, 2016).

Larsen et al.~(2013) designed a tool to prioritize the placement of new
cycling facilities in Montreal, Canada. The tool used survey, vehicle
collision, and origin-destination data. To create the index, they first
identified a set of relevant indicators for prioritising locations for
infrastructure investment. Then, they imposed a grid and aggregated the
indicators spatially, associating them with grid's cells. Finally, they
combined the aggregated grid cell indicators into the prioritization
index. The tool generated heatmaps recommending cycling infrastructure
upgrades in specific locations.

Zhang et al.~(2014) used a regression model to examine the effect of
socioeconomic factors on willingness to cycle in Belo Horizonte City,
Brazil, while considering individual heterogeneity. To develop the
model, they employed a bicycle use survey conducted in Belo Horizonte.
The results indicated a positive association between willingness to
cycle and middle income, people renting apartments, and higher
education, and a negatively association with distance to work and car
use. Subsequently, using the calibrated model along with census and OD
data, they estimated the willingness to commute cycling for the entire
city.

Phillips and Range (2017) created the Cycling Potential Tool (CPT) for
Scotland to provide evidence-based insights for decision-making
regarding local cycling infrastructure in Scotland. The CPT is a GIS
toolbox consisting of two main modules: the base environmental module
and the quality of service module. The base environmental module is
based on eight weighted factors (population density, hilliness, physical
barriers, access to services, existing cycling mode share, distance to
work and school, and road speed) and produces heatmaps for each factor
as well as a combined score. The quality of service module evaluates the
network within each area of interest based on eight factors (surface
condition, adjacent cyclists, comfort factor, conflict, distance between
junctions, slope, access to services, and origin/destination).

The Propensity to Cycle Tool (2017) was initially based on current and
future potential uptake of cycling for single stage travel to work at
desire line, zone, route, and route network levels (Lovelace et
al.~2017). Extensions of the PCT approach have included estimation of
benefits at the individual level (Woodcock et al., 2018), addition of
travel to school network (Goodman et al., 2019), and improved modeling
of impacts on health, environmental and distributional outcomes
(Woodcock et al., 2021). Initially developed just for England, the PCT
was extended to cover all of Wales (for commuter data only) in 2018. The
PCT approach has been applied in other countries, including Ireland (the
topic of this paper), Scotland, and Portugal. In Portugal the `biclaR'
project, based on methods underlying the PCT, has been developed and
deployed for the Lisbon metro region. The resulting evidence is
available in an interactive web application hosted at
biclar.tmlmobilidade.pt (Félix et al., 2022). biclaR includes estimates
of impacts, using the World Health Organisation (WHO) `HEAT for Cycling'
tool and an `intermodality' scenario that combines cycling with
currently available public transit options based on General Transit Feed
Specification (GTFS) data. An important feature of the PCT is that it is
open source and publicly available (at www.pct.bike), allowing its use
by all stakeholders in the transport planning process.

Silva et al.~(2021) designed the Gross Potential for Cycling (GPC) tool
to identify key areas for the implementation of cycling infrastructure
and other cycling measures. The GPC uses two sets of indicators:
population-based indicators (age, potential demand density, employment
density and motorization rate) and area-based indicators (accessibility,
education, public transport, connectivity, land use mix and relative
performance). These indicators are calculated considering topography,
road hierarchy, and average congestion and presented on a scale from 5
to 10. They are then combined into an overall score, weighted by their
impact on cycling according to the literature. The GPC is publicly
available (at https://boost.up.pt/en/ferramentas/gpc) and is specially
designed for local council-level practitioners. Its practical value has
been assessed through a series of experiential workshops involving
planning practitioners from various Portuguese contexts (Silva et al.,
2022).

The methods used to generate the evidence present in the CRUSE Tool,
described in the next section, build on the PCT and seek to overcome
three of the PCT methodological limitations: 1) low resolution of data,
with routes starting and ending in administrative zone centroids, 2)
limited coverage of trip purposes beyond travel to work and school, and
3) a web interface that was not user-friendly or intuitive.

@@
